A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-03 latex/main.tex
Tidy up supp typesetting

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, using small numbers of widely diverged
genomes. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of genomes sampled from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods [citation?]
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly limited to
visualization, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

The global
surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer amount of data 
overwhelmed even approximate distance-matrix based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February 2022 about 80 thousand samples
per day  were submitted to GISAID\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October 2025.
Regularly reinferring trees de-novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real-time~\cite{McBroome2021}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is an important force
in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present
in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts \cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES ~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
We found that they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and a previous study \cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of
the ARG and UShER tree closely match by generating tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree reasonably well with
NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Document \ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023},
but are not incorporated into UShER trees and
therefore must be accounted for separately.
Because deletions are difficult to distinguish
% SHZ: Sequencing errors or misleading alignment.
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using \textit{post hoc} parsimony at 75 sites
involved in deletions with a frequency of
$\geq 1\%$ reported by Li et al.\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example,
the three defining deletions of Alpha \cite{RambautAlpha2020}
(11288:9, 21765:6, 21991:3)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hot spot \cite{McCarthy2021}),
for example,
the 21765:6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry \cite{Meng2021}, and
the 21991:3 deletion
which is associated with antibody escape \cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombinant parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two ``Group A'' recombinant samples reported by Jackson et al.\cite{Jackson2021} are shown.
(B) Subgraph around the recombination node associated with a 
group of ``nested'' Pango X lineages: we use the 
label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis.
The first recombinant lineage XA was designated in 2021
based on the analysis of Jackson et al.\cite{Jackson2021},
which established that recombination with onward transmission had occurred.
Although the process of incorporating recombinant lineages
in the Pango nomenclature system is well defined,
the identification of such events is much more subjective,
requiring the synthesis of multiple lines of evidence and
often the manual inspection of phylogenies and sequence alignments
by volunteers~\cite{Smith2023tracking}.
% Debatable whether this adds or substracts from narrative
As of pango-designation version v1.33.1 (2025-03-31),
a total of 146 distinct Pango X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
We designate these as Type I events, where the ARG reflects the
hierarchical structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt \cite{Alfonsi2024}
and CovRecomb \cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
% NOTE: deleted the "sister" bit because it's confusing when
% reading the document at this level. Keep the detailed hyptothese
% to the deep dives.
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages (Document \ref{sec:dd_xq}). 
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B, Document \ref{sec:dd_xx}).

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document \ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized \cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (Document \ref{sec:dd_xb}).

@@

The location of a recombination breakpoint is generally
impossible to pinpoint exactly, and instead an interval
containing the breakpoint can be inferred by
examining the parental and recombinant sequences.
Some methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
and approaches (e.g., Pango designation) infer this
as the genomic distance between lineage-associated
mutations that differentiate the two parental \emph{lineages}.
Similarly, methods based on probabilistic representations of
mutational profiles \cite{Varabyou2021rapid}
can only be as precise as examination of the two
parent lineages allows.
However, genealogical methods (such as sc2ts and RIPPLES) are
based on identifying exact parental \emph{sequences},
so are in principle maximally precise.
The breakpoint interval of a recombinant is bounded by
two positions: a left position where the recombinant
matches the left parent sequence but not the right, and
a right position where the recombinant matches the right
parent sequence but not the left. 
How well the time and position of a recombination event
can be inferred thus depends on the divergence between the
parent sequences (measured as time to their MRCA).

The distribution of breakpoint intervals of
the QC-passing recombination events
shows an overall enrichment of breakpoints
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies \cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
making recombination events more likely to be detectable, and 
allowing breakpoint intervals to be estimated more precisely.
We found the median breakpoint interval length to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably larger (up to $\sim$54\% of the genome).
As expected, we found a negative relationship between
the length of a breakpoint interval and
divergence between recombinant parents
(Figure~\ref{fig:recombinant_intervals}D),
which follows theoretical expectation (STAR Methods).
Breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

@@

We then asked which single measure best explains
the number of recombination events inferred each week,
considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
We included the squared number of cases each week
thinking that it might better predict co-infections.
The best single predictor of recombination events involved
the product of cases and expected heterozygosity ($n H$),
with an adjusted $R^2$ of 23.7\%.
This measure of case diversity ($n H$)
explained 49\% more of the variation in recombination events per week
than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
While consistent with the hypotheses
that recombination events occur more often when infection rates are common
and are detected more often when viral diversity is high,
we note several sources of uncertainty.
Global case numbers were underreported,
particularly during the Delta and Omicron waves \cite{Wang2025}.
Furthermore, there may be substantial uncertainty in the timing of a recombination event,
particularly if these occur in chronically infected individuals
(as reported by \cite{Burel2022})
with substantial delays before onward cases are detected.
Finally, the filters that we applied removed all
but the most strongly supported recombination events,
although similar conclusions were obtained
when all 855 recombinants were considered (STAR Methods).

@@

https://github.com/jeromekelleher/sc2ts-paper/issues/287

@@

>>>>

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, using small numbers of widely diverged
genomes. In this context, recombination is a nuisance which must be 
carefully accounted for~\cite{Rannala2025recombination},
affecting the inference of species trees~\cite{Posada2002effect}
and biasing downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}.
Modern pathogen datasets are not sparsely sampled from different species,
but consist of tens to hundreds of thousands of genomes sampled from 
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to perform species-level tree
inference~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of these datasets are typically built with
simpler distance methods % citation for this? 
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations,
but also because of the complexity of modeling such datasets.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly limited to
visualization, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

The global
surveillance data collected during the COVID-19 pandemic presents fundamental challenges to existing approaches
in two ways. Firstly, the sheer amount of data 
overwhelmed even approximate distance-matrix based methods.
% Accessed GISAID/EpiCov on 08 Oct., 2025
At the peak in February 2022 about 80 thousand samples
per day  were submitted to GISAID\cite{Shu2017},
accumulating to a total of over 17.5 million samples
as of October 2025.
Regularly reinferring trees de-novo from such rapidly
growing datasets is untenable.
The UShER\cite{Turakhia2021} method is based on adding
samples to an existing tree using maximum parsimony,
with periodic rearrangements to counter the inherent 
greediness of the approach\cite{DeMaio2023}.
The benefits of this simple model and the value of
a comprehensive genealogy updated in real-time~\cite{McBroome2021,Kramer2023}
were conclusively demonstrated, as the resource
quickly became an indispensable element
of the pandemic response\cite{Hinrichs2024}.
The second challenge posed is that recombination is an important force
in SARS-CoV-2 evolution \cite{Jackson2021,VanInsberghe2021,Focosi2022,
Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present
in the genealogy of the global population \cite{Roemer2023,Shiraz2023enhanced}.
Although several methods have been proposed to \emph{detect}
recombinant samples~\cite{
Varabyou2021rapid,% Bolotie
Turakhia2022,%RIPPLES
Zhou2023virusrecom,% CovRecomb
Alfonsi2024,%RecombinHunt 
Li2024CovRecomb%CovRecomb 
},
this information is not incorporated into the phylogenies used 
in innumerable studies~\cite{McBroome2021,Hadfield2018},
and recombination continues to be treated as a 
nuisance factor.

@@

Here we present sc2ts, a new method to infer ARGs for SARS-CoV-2  
that incrementally adds samples over time in a similar manner 
to UShER, but also automatically detects and incorporates recombination
in real-time. Based on the Viridian dataset, a freely available  
resource purged of various systematic errors and artifacts~\cite{Hunt2024},
we infer an ARG with 2.48 million whole genome sequences.
We show that this ARG is highly congruent with the UShER phylogeny built 
from the same dataset as well as the Pango nomenclature system~\cite{Rambaut2020},
and also accurately captures evolutionary features such as mutational 
spectra~\cite{Bloom2023} and deletions\cite{Li2023indels}.
We show that sc2ts automatically and accurately detects details about
known recombinant lineages that were painstakingly collated in a community 
process, and provides novel insights into the relationships between these
recombinant lineages. We compare the recombination events detected by 
sc2ts with the state-of-the-art combination of UShER and 
RIPPLES~\cite{Turakhia2022}, showing that sc2ts is more sensitive,
detecting more of the well-characterized recombination events 
curated by the Pango lineage designation community.
The output of sc2ts---incorporating
genealogical relationships, point mutations, deletions, and 
recombination---provides a basis for the study of SARS-CoV-2
that systematically accounts for all of these evolutionary forces.
This ARG, coupled with the rich suite of supporting software
in the tskit~\cite{Ralph2020} and VCF-Zarr~\cite{Czech2025} ecosystems,
enables pandemic-scale analysis on a laptop. 
These resources
will help us to learn from the vast troves of data collected during the
pandemic, and to prepare for the next.

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
We found that they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of
the ARG and UShER tree closely match by generating tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree reasonably well with
NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Document~\ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

Deletions are an important part of SARS-CoV-2 evolution~\cite{Jeronimo2023},
but are not incorporated into UShER trees and
therefore must be accounted for separately.
Because deletions are difficult to distinguish
% SHZ: Sequencing errors or misleading alignment.
from bioinformatics errors
and their multi-base nature presents challenges
to the Markovian structure of the LS HMM,
we masked out all gap characters as missing data during primary inference.
To approximate the effects of the most important deletions into the ARG,
we remapped data using \textit{post hoc} parsimony at 75 sites
involved in deletions with a frequency of
$\geq 1\%$ reported by Li et al.~\cite{Li2023indels}
during post-processing (STAR Methods).
Although sites are treated independently in this approach,
runs of mutations to the gap character at adjacent sites
correctly identified many known
high-frequency deletions (Table~\ref{tab:major_dels}) and
recurrent deletions (Table~\ref{tab:rec_dels}).
For example,
the three defining deletions of Alpha~\cite{RambautAlpha2020}
(11288:9, 21765:6, 21991:3)
are correctly mapped above the Alpha origin in the ARG
(Figure~\ref{fig:subgraph_alpha}).
The ARG captures known recurrent deletions
in the N-terminal domain of the S1 unit of Spike
(which is a deletion hot spot~\cite{McCarthy2021}),
for example,
the 21765:6 deletion ($\Delta$H69/V70),
which causes S-Gene Target Failure \cite{Walker2021} and
is associated with increased efficiency in cell entry~\cite{Meng2021}, and
the 21991:3 deletion
which is associated with antibody escape~\cite{McCarthy2021}.
See Document~\ref{sec:supplement_dels} for further analysis.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombinant parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two ``Group A'' recombinant samples
reported by Jackson et al.~\cite{Jackson2021} are shown.
(B) Subgraph around the recombination node associated with a 
group of ``nested'' Pango X lineages: we use the 
label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

% " Our nomenclature system should be referred to as the ‘Pango’ nomenclature
% system ... ". It should not be referred to as the ‘Rambaut et al.’ or
% ‘Pangolin’ nomenclature system. "
Part of the success of the Pango nomenclature system~\cite{Rambaut2020} 
can be attributed to the clarity of the rules under which new lineages 
are designated and their explicitly phylogenetic basis.
The first recombinant lineage XA was designated in 2021
based on the analysis of Jackson et al.~\cite{Jackson2021},
which established that recombination with onward transmission had occurred.
Although the process of incorporating recombinant lineages
in the Pango nomenclature system is well defined,
the identification of such events is much more subjective,
requiring the synthesis of multiple lines of evidence and
often the manual inspection of phylogenies and sequence alignments
by volunteers~\cite{Smith2023tracking}.
% Debatable whether this adds or substracts from narrative
As of pango-designation version v1.33.1 (2025-03-31),
a total of 146 distinct Pango X lineages have been designated.
%$ cut -f1 ../arg_postprocessing/pangonet_data/lineage_notes.txt | grep '^X' | grep -v '\.' | wc -l 
% 146

By automatically detecting and incorporating recombination events into the ARG,
sc2ts provides us with a systematic means of
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarizes
these results for the 44 Pango X lineages present in the ARG
(XAC and XAD have been omitted, and XM partially represented,
corresponding to 23 omitted samples in total;
see Document~\ref{sec:supplement_x_pango_lineages}).
For 16 Pango X lineages, the oldest node
that has been assigned to the focal Pango lineage is a recombination node, and
that node is the ancestor of all samples
assigned to that (or a descendant) Pango lineage.
We designate these as Type I events, where the ARG reflects the
hierarchical structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent,
with similar performance to
the RecombinHunt~\cite{Alfonsi2024}
and CovRecomb~\cite{Li2024CovRecomb} detection methods
(STAR Methods; Table~\ref{tab:method_concordance}; see also
Document~\ref{sec:jackson_recombs} for analysis of the 
Jackson et al.\ recombinants).
Of the Type I events, 13 are monophyletic for the focal Pango lineage
(i.e., the descendants are entirely from that lineage);
Figure~\ref{fig:subgraphs}A shows XA as an example.

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
% NOTE: deleted the "sister" bit because it's confusing when
% reading the document at this level. Keep the detailed hyptothese
% to the deep dives.
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages (Document~\ref{sec:dd_xq}). 
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B, Document~\ref{sec:dd_xx}).

The final class of events (Type IV) in Table~\ref{tab:pango_x_lineages}
corresponds to nodes that are not associated with a recombination event.
These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence
to suggest recombination (Document~\ref{sec:dd_xn}).
XB, on the other hand,
has been thoroughly characterized~\cite{Gutierrez2022} and
may have misclassified by sc2ts due to limitations in
handling deletions or matching against long branches (Document \ref{sec:dd_xb}).

@@

The location of a recombination breakpoint is generally
impossible to pinpoint exactly, and instead an interval
containing the breakpoint can be inferred by
examining the parental and recombinant sequences.
Some methods~\cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
and approaches (e.g., Pango designation) infer this
as the genomic distance between lineage-associated
mutations that differentiate the two parental \emph{lineages}.
Similarly, methods based on probabilistic representations of
mutational profiles \cite{Varabyou2021rapid}
can only be as precise as examination of the two
parent lineages allows.
However, genealogical methods (such as sc2ts and RIPPLES) are
based on identifying exact parental \emph{sequences},
so are in principle maximally precise.
The breakpoint interval of a recombinant is bounded by
two positions: a left position where the recombinant
matches the left parent sequence but not the right, and
a right position where the recombinant matches the right
parent sequence but not the left. 
How well the time and position of a recombination event
can be inferred thus depends on the divergence between the
parent sequences (measured as time to their MRCA).

The distribution of breakpoint intervals of
the QC-passing recombination events
shows an overall enrichment of breakpoints
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies~\cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
making recombination events more likely to be detectable, and 
allowing breakpoint intervals to be estimated more precisely.
We found the median breakpoint interval length to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably larger (up to $\sim$54\% of the genome).
As expected, we found a negative relationship between
the length of a breakpoint interval and
divergence between recombinant parents
(Figure~\ref{fig:recombinant_intervals}D),
which follows theoretical expectation (STAR Methods).
Breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

@@

We then asked which single measure best explains
the number of recombination events inferred each week,
considering $n$, $H$, $n H$, $n^2$, $n^2 H$.
We included the squared number of cases each week
thinking that it might better predict co-infections.
The best single predictor of recombination events involved
the product of cases and expected heterozygosity ($n H$),
with an adjusted $R^2$ of 23.7\%.
This measure of case diversity ($n H$)
explained 49\% more of the variation in recombination events per week
than the next best predictor ($H$, adjusted $R^2$ of 15.9\%).
While consistent with the hypotheses
that recombination events occur more often when infection rates are common
and are detected more often when viral diversity is high,
we note several sources of uncertainty.
Global case numbers were underreported,
particularly during the Delta and Omicron waves~\cite{Wang2025}.
Furthermore, there may be substantial uncertainty in the timing of a recombination event,
particularly if these occur in chronically infected individuals
(as reported by \cite{Burel2022})
with substantial delays before onward cases are detected.
Finally, the filters that we applied removed all
but the most strongly supported recombination events,
although similar conclusions were obtained
when all 855 recombinants were considered (STAR Methods).

@@

@@

\scghdiscussion{782}
